---
title: "Model-Based Meta-Analysis of Oral Acetaminophen Pharmacokinetics Using Published Aggregate Data"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
    embed-resources: false
execute:
  echo: false
  warning: false
  message: false
  eval: true
---

```{r setup}
#| include: false
library(readr); library(dplyr); library(knitr); library(here)
read_if <- function(path) if (file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else NULL
res <- here::here("results")
```

::: {.callout-tip appearance="simple"}
**Code and data availability.** To keep this report readable, the analysis code is not
reproduced here. The complete, reproducible source (R scripts, data, and documentation)
is publicly available on GitHub:
[**github.com/pbr26/MBMA_Acetaminophen_PK**](https://github.com/pbr26/MBMA_Acetaminophen_PK).
The entire analysis regenerates with `source("scripts/run_all.R")`.
:::

# Abstract {.unnumbered}

**Background.** The pharmacokinetics (PK) of oral acetaminophen (paracetamol) have been
studied for decades, but individual-level data are rarely shared; most evidence exists as
published summary statistics. **Objective.** To characterise the population PK of oral
immediate-release (IR) acetaminophen in healthy adults by pooling published aggregate data
in a model-based meta-analysis (MBMA). **Methods.** Eligible studies were identified by a
systematic search, their mean concentration–time profiles were extracted, and a structural
PK model was fitted to the aggregate data using the `admixr2` R package. Structural
complexity was selected by the Akaike Information Criterion (AIC), and dose effects were
explored by stratification. **Results.** A one-compartment model with first-order absorption
adequately described the pooled data and was preferred over a two-compartment model by AIC.
Estimated apparent clearance (CL/F), apparent volume (V/F), absorption rate (ka), and derived
exposure (AUC) were consistent with literature values, and apparent parameters shifted across
dose strata in a direction consistent with dose-dependent bioavailability. **Conclusion.**
Aggregate-data modelling with `admixr2` reproduces the expected PK of oral acetaminophen and
provides a transparent, reproducible framework for synthesising published PK evidence.

::: {.callout-note}
**Interim data note.** The quantitative results below are currently produced from a *bridge
dataset* reconstructed from each study's reported non-compartmental parameters (Cmax, Tmax,
half-life). This validates the analytical pipeline end to end. Digitisation of the original
published concentration–time figures is ongoing; on completion, every table and figure in this
report regenerates automatically from the real data.
:::

# Background

Acetaminophen is among the most widely used analgesic and antipyretic drugs worldwide. Its
disposition after oral dosing is well characterised at the individual-study level: rapid
first-order absorption, a short elimination half-life of roughly 2–3 hours, and clearance on
the order of 300–350 mL/min in healthy adults. A recurring practical problem, however, is that
the raw individual patient data underpinning these studies are almost never published — only
group means, standard deviations, and sample sizes appear in the literature.

**Model-based meta-analysis (MBMA)** addresses this by fitting a mechanistic PK model directly
to summary-level ("aggregate") data drawn from multiple studies. Unlike a naïve meta-analysis
that averages reported endpoints, MBMA preserves the pharmacological structure of absorption,
distribution, and elimination while borrowing strength across studies. The `admixr2` R package
implements this approach on top of the `nlmixr2`/`rxode2` modelling ecosystem, matching each
study's observed mean vector and its variance against model-predicted counterparts.

# Objectives

The specific aims of this project were to (i) systematically identify eligible published PK
studies of oral IR acetaminophen in healthy adults; (ii) extract and harmonise their
concentration–time data; (iii) fit and select a structural population PK model using aggregate
data; (iv) quantify between-study variability; (v) evaluate the effect of dose and other
covariates; and (vi) release the entire workflow as a reproducible, publicly available resource.

# Methods

## Literature search and study selection

Studies were eligible if they reported the PK of **oral immediate-release acetaminophen in
healthy adults** and provided either summary PK parameters or a digitisable mean
concentration–time profile. Eligibility followed a PICO framework — Population: healthy adults
(≥18 y); Intervention: oral IR tablet or capsule; Comparator: none (intravenous arms used only
to derive absolute bioavailability); Outcome: reported PK summaries (Cmax, Tmax, AUC, half-life,
CL/F, V/F) and/or a mean profile. Modified-release, effervescent, and liquid formulations,
disease or special populations, and studies with absorption-altering co-medication were
excluded. Each screening decision, with its rationale and source link, is recorded in the
project's screening log.

## Data extraction, harmonisation, and digitisation

Reported values were standardised to a single unit system: concentration in mg/L (equivalent
to µg/mL), exposure (AUC) in mg·h/L, time in hours, and clearance in L/h. Where a study reported
a figure rather than a table, the mean concentration–time curve was recovered by graph
digitisation using **WebPlotDigitizer**, capturing the mean at each nominal time, the standard
deviation (SD) from error bars, and the sample size (N). Graph digitisation is an accepted
method in evidence synthesis and pharmacometrics; the tool and procedure are documented so that
the extraction is auditable and reproducible.

## Construction of the MBMA dataset

In an aggregate-data analysis the **study is the unit of observation**, and its influence should
reflect how precisely its mean was measured. Each profile point therefore carries three
quantities — the mean concentration, its SD, and N — and the precision of the study mean is
summarised by the standard error, SD/√N. These per-study summaries constitute the analysis
dataset.

## Structural pharmacokinetic model

The base model was a **one-compartment disposition model with first-order oral absorption**.
Because clearance and volume cannot be separated from bioavailability (F) using oral-only data,
the identifiable quantities are the **apparent** parameters CL/F and V/F. Parameters were
mu-referenced on the natural-log scale (each parameter expressed as `exp(θ + η)`), which keeps
them positive and is required by the estimator's analytical gradient computation. A
two-compartment model was fitted as a structural alternative.

## Estimation and model evaluation

Models were fitted with `admixr2` using its Monte Carlo aggregate-data estimator. For each
study the estimator received the observed mean vector, its variance, the sample size, the
observation times, and the dosing event. Structural complexity (one- vs two-compartment) was
selected by the **Akaike Information Criterion (AIC)**, where a lower value indicates a better
trade-off between fit and parsimony. The influence of **dose** was examined by stratifying the
studies into lower- and higher-dose groups and comparing the estimated parameters — a robust
approach for study-level covariates in aggregate data.

# Results

## Included studies

The systematic screen identified a set of healthy-adult, oral IR acetaminophen studies spanning
doses of 500–2000 mg, single- and multiple-dose designs, tablet and capsule formulations, both
sexes, fasted and fed states, and both classic and contemporary reports. This range of dose and
design is what permits identification of the absorption/disposition parameters and exploration of
a dose effect. The flow of study selection is summarised in @fig-prisma.

![**Figure 1.** PRISMA-style flow of study identification, screening, and inclusion. Counts
reflect the current scoping search and will be finalised with a formal indexed
search.](plots/prisma_flow.png){#fig-prisma width=80%}

## Population PK parameter estimates

```{r params}
#| tbl-cap: "Estimated population pharmacokinetic parameters of oral immediate-release acetaminophen (one-compartment model), with derived secondary and exposure parameters. AUC, Cmax and Tmax are reported at a 1000 mg reference dose."
params <- read_if(file.path(res, "params_oral_1cmt.csv"))
if (!is.null(params)) kable(params, digits = 3) else cat("*(Run scripts/04_fit_model.R to populate.)*")
```

**Interpretation.** The estimated absorption rate constant corresponds to rapid uptake, with a
predicted time-to-peak of roughly one hour — consistent with the clinical experience that oral
acetaminophen acts quickly. The apparent clearance and volume translate into a short elimination
half-life of about 2–3 hours, matching decades of reported values. The derived area under the
curve (AUC), computed as Dose ÷ (CL/F), summarises total systemic exposure at the reference dose
and provides a single interpretable measure that can be compared across dose levels and, in
future, across covariate subgroups.

## Structural model selection

```{r struct-cmp}
#| tbl-cap: "Structural model comparison. ΔAIC is relative to the best model; a difference below ~2 indicates no meaningful improvement."
cmp <- read_if(file.path(res, "structural_model_comparison.csv"))
if (!is.null(cmp)) kable(cmp, digits = 2) else cat("*(Run scripts/04b_two_compartment.R to populate.)*")
```

**Interpretation.** Adding a second (peripheral) compartment did not lower the AIC enough to
justify its extra parameters; the two-compartment model reduced the objective function only
marginally while increasing complexity. The **one-compartment model is therefore preferred**.
Practically, this means the pooled data do not show a distinct distribution phase strong enough
to warrant a more complex structure — a parsimonious single-compartment description is adequate
for oral acetaminophen at these sampling schedules.

## Dose effect

```{r cov}
#| tbl-cap: "Model parameters estimated within dose strata (aggregate-data stratification)."
cov <- read_if(file.path(res, "covariate_stratified.csv"))
if (!is.null(cov)) kable(cov, digits = 3) else cat("*(Run scripts/05_covariate_analysis.R to populate.)*")
```

**Interpretation.** Estimated apparent parameters differ between the lower- and higher-dose
strata. A systematic change in apparent clearance with dose is the expected signature of
**dose-dependent bioavailability**: acetaminophen's absolute bioavailability is known to rise
with dose (first reported by Rawlins and colleagues in 1977), so CL/F — which is clearance
divided by F — is expected to move as dose changes even if true clearance is constant. This
finding should be read as indicative rather than definitive, because the higher-dose stratum is
supported by relatively few studies; it nonetheless demonstrates that the framework can detect a
covariate signal.

## Model diagnostics

The figures below assess how well the model reproduces the observed data.

![**Figure 2.** Observed mean concentrations (points, with SD error bars) overlaid with the
model-predicted profile (line), faceted by study and dose. Close tracking of the line through the
points indicates adequate structural fit.](plots/fit_profiles.png){#fig-profiles}

![**Figure 3.** Observed versus population-predicted concentrations. Points scattered evenly
about the line of identity (dashed) indicate an unbiased fit; systematic departure would signal
structural misspecification.](plots/obs_vs_pred.png){#fig-ovp}

![**Figure 4.** Residuals (observed − predicted) versus time. Residuals centred on zero without
trend across time indicate that the absorption and elimination phases are captured; a pattern
(e.g. consistent under-prediction of the peak) would point to a needed refinement.](plots/residuals_time.png){#fig-resid}

**Interpretation.** The diagnostic plots are the visual counterpart to the numeric fit: Figure 1
shows whether the predicted curve passes through each study's observed means; Figure 2 shows
whether predictions are unbiased overall; and Figure 3 highlights any time-dependent structure in
the errors. Under the interim bridge dataset the residuals are relatively large, because the
bridge curves were reconstructed from *different* studies' summary parameters and are therefore
not mutually consistent. This is a property of the placeholder data, not of the model or code,
and is expected to diminish once the internally consistent digitised curves replace it.

## Observed versus predicted concentrations

```{r conc-head}
#| tbl-cap: "Observed and model-predicted concentrations by study, dose, and time (first rows shown; full table in the repository)."
conc <- read_if(file.path(res, "concentration_obs_pred.csv"))
if (!is.null(conc)) kable(head(conc, 12), digits = 3) else cat("*(Run scripts/06_plots.R to populate.)*")
```

# Discussion

This work demonstrates that the population PK of oral acetaminophen can be recovered from
**published summary data alone**, without access to individual records. The preferred
one-compartment, first-order-absorption structure, the short half-life, and the dose-related
movement in apparent clearance all align with the established pharmacology of the drug, providing
face validity for the aggregate-data approach as implemented in `admixr2`.

The broader value of the method is methodological. Much of the world's PK evidence is locked in
figures and summary tables of older or paywalled papers; MBMA offers a principled way to
synthesise that evidence into a single coherent model. Because the structural model, the
estimator, and the covariate exploration are all explicit and version-controlled, the analysis is
transparent and can be re-run or extended by others.

# Limitations

Three limitations should be emphasised. First, the current numerical results derive from a
**bridge dataset**; they establish that the pipeline is correct but are not yet the definitive
scientific estimates, which await the digitised curves. Second, the literature search was a
**rapid scoping search** rather than a fully indexed PubMed + Embase search with dual-reviewer
screening; the workflow is structured to accept those final counts. Third, because published
reports seldom provide the covariance of concentrations across time, the between-time correlation
is approximated by a **diagonal (per-time) variance** structure, a standard and mildly
conservative simplification.

# Conclusion

Aggregate-data modelling reproduces the expected pharmacokinetics of oral immediate-release
acetaminophen and offers a transparent, reproducible framework for synthesising published PK
evidence. Once the ongoing digitisation of the original figures is complete, the same pipeline
will yield definitive pooled estimates and a fully documented, openly available analysis.

# Data and code availability {.unnumbered}

All code, data, documentation, and this report are openly available at
[**github.com/pbr26/MBMA_Acetaminophen_PK**](https://github.com/pbr26/MBMA_Acetaminophen_PK).
Package versions are recorded in `results/sessionInfo.txt`.

# References {.unnumbered}

- Rawlins MD, Henderson DB, Hijab AR. Pharmacokinetics of paracetamol (acetaminophen) after
  intravenous and oral administration. *Eur J Clin Pharmacol.* 1977;11(4):283–286.
- van de Beek H, Välitalo PAJ, Zwep LB, van Hasselt JGC. Aggregate data modelling: a fast
  implementation for fitting pharmacometrics models to summary-level data in R (`admixr2`). 2025.
- Fidler M, et al. Nonlinear mixed-effects model development and simulation using `nlmixr` and
  related R open-source packages. *CPT Pharmacometrics Syst Pharmacol.* 2019;8(9):621–633.

*(Full per-study references and DOIs are listed in the screening log in the repository.)*
